---
title: Final Prediction
author: Victor Bowker
date: '2024-10-27'
slug: final-prediction
categories: []
tags: []
---

# Disregard - Incomplete

(1) model formula (or procedure for obtaining prediction), 
(2) model description and justification, 
(3) coefficients (if using singular regression) and/or weights (if using ensemble of different models) and/or feature importance (if using machine learning)
(4) interpretation of coefficients, feature importance, and/or justification of weights, 
(5) model validation (recommended to include both in-sample and out-of-sample performance unless it is impossible due to the characteristics of model and related data availability), 
(6) uncertainty around prediction (e.g. predictive interval)
(7) graphic(s) showing your prediction


# Welcome, Again

Can you believe it? This is my final post before the 2024 Presidential General Election, where either Former President Donald Trump, or Vice President Kamala Harris, will be elected the next President of the United States. I think




```{r, include=FALSE, echo=FALSE, warning=FALSE, message=FALSE}
options(repos = c(CRAN = "https://cloud.r-project.org"))

```

```{r, include=FALSE, echo=FALSE, warning=FALSE, message=FALSE}
install.packages("patchwork")
library(patchwork)
library(broom)
library(geofacet)
library(ggpubr)
library(ggthemes)
library(haven)
library(DT)
library(maps)
library(mgcv)
library(mgcViz)
library(RColorBrewer)
library(scales)
library(sf)
library(spData)
library(stargazer)
library(tidygeocoder)
library(tidyverse)
library(tigris)
library(tmap)
library(tmaptools)
library(viridis)
library(tidyverse)
library(gt)
install.packages("readxl")
library(readxl)
```

```{r, include=FALSE, echo=FALSE, warning=FALSE, message=FALSE}
# Read popular vote datasets. 
d_popvote <- read_csv("popvote_1948_2020.csv")
d_state_popvote <- read_csv("state_popvote_1948_2020.csv")
states <- map_data("state") |>
  rename(state = region)
states$state <- str_to_title(states$state)
d_state_popvote[d_state_popvote$state == "District of Columbia",]$state <- "District Of Columbia"
```

```{r, include=FALSE, echo=FALSE, warning=FALSE, message=FALSE}
# Read elector distribution dataset. 
d_ec <- read_csv("corrected_ec_1948_2024.csv")
d_ec <- d_ec |>
  filter(year == 2024)
library(readxl)
expert <- read_excel("/Users/victorbowker/Fall 2024/election-blog/content/post/2024-10-26-blog-post-8/expert_2024-2.xlsx")

```

```{r, include=FALSE, echo=FALSE, warning=FALSE, message=FALSE}
# Read polling data. 
d_polls <- read_csv("national_polls_1968-2024.csv")
d_state_polls <- read_csv("state_polls_1968-2024.csv")
```

```{r, include=FALSE, echo=FALSE, warning=FALSE, message=FALSE}
# Read turnout data. 
d_turnout <- read_csv("state_turnout_1980_2022.csv")
```

```{r, include=FALSE, echo=FALSE, warning=FALSE, message=FALSE}
# Read county turnout. 
d_county_turnout <- read_csv("county_turnout.csv")
```

```{r, include=FALSE, echo=FALSE, warning=FALSE, message=FALSE}
# Read state-level demographics.
d_state_demog <- read_csv("demographics.csv")
```

```{r, include=FALSE, echo=FALSE, warning=FALSE, message=FALSE}
# Read county demographics. 
d_county_demog <- read_csv("county_demographics.csv")
```

```{r, include=FALSE, echo=FALSE, warning=FALSE, message=FALSE}
# Read campaign events datasets. 
d_campaign_events <- read_csv("campaigns_2016_2024.csv")[,-1]
```


```{r, echo=FALSE, warning=FALSE, message=FALSE, include=FALSE}

state_abb_xwalk <- d_state_popvote |>
  mutate(state_abb = state.abb[match(d_state_popvote$state, state.name)]) |> 
  select(state, state_abb) |> 
  distinct() 
state_abb_xwalk[51,]$state <- "District Of Columbia"
state_abb_xwalk[51,]$state_abb <- "DC"
```



```{r}
d_campaign_events <- read_csv("campaign_events_geocoded.csv")
d_campaign_events$party[d_campaign_events$candidate %in% c("Trump / Pence", "Trump", "Pence", "Trump/Pence", "Vance")] <- "REP"
d_campaign_events$party[d_campaign_events$candidate %in% c("Biden / Harris", "Biden", "Harris", "Biden/Harris", "Walz", "Kaine", "Clinton", "Clinton / Kaine")] <- "DEM"

d_campaign_events
```

```{r, echo=FALSE, warning=FALSE, message=FALSE, include=FALSE}
kaitlyn_dataset <- 
  d_campaign_events |> 
  filter(party == "DEM") |>
  left_join(state_abb_xwalk, by = "state") |> 
  left_join(d_state_popvote, by = c("state", "year")) |>
  group_by(state, year) |>
  summarize(fun_campaign_events = n()) 
```

```{r, echo=FALSE, warning=FALSE, message=FALSE, include=FALSE}
library(dplyr)
library(tidyr)
library(lubridate)

model44 <- d_state_polls |>
  select(state, year, weeks_left, poll_date, poll_support, party) |>
  filter(month(as.Date(poll_date)) %in% c(9, 10), party == "DEM", year >= 2016) |>
  filter(year < 2024) |>
  group_by(year, state, month = month(as.Date(poll_date))) |>
  summarize(poll_support = round(weighted.mean(poll_support, weeks_left, na.rm = TRUE), 3), .groups = "drop") |>
  pivot_wider(names_from = month, 
              values_from = poll_support, 
              names_prefix = "month_") |>
  left_join(d_state_popvote, by = c("year", "state")) |>
  select(year,state, month_9, month_10, D_pv2p) |>
  rename(September_Results = month_9, 
         October_Results = month_10) |>
  left_join(kaitlyn_dataset, by = c("state", "year")) |>
  mutate(across(everything(), ~ replace_na(., 0))) 


```

```{r, echo=FALSE, warning=FALSE, include=FALSE, message=FALSE}
reg44 <- lm(D_pv2p ~ September_Results + October_Results + fun_campaign_events, 
                data = model44)
```



```{r, echo=FALSE, warning=FALSE, message=FALSE, include=FALSE}
tidy_model44 <- tidy(reg44)
```

```{r, echo=FALSE, warning=FALSE, message=FALSE}
datatable(tidy_model44, 
          options = list(pageLength = 4), 
          caption = "Updated Regression Results") 
```

```{r, echo=FALSE, warning=FALSE, message=FALSE, include=FALSE}

prediction_data <- d_state_polls |>
  select(state, year, weeks_left, poll_date, poll_support, party) |>
  filter(month(as.Date(poll_date)) %in% c(9, 10), party == "DEM", year == 2024) |>
  group_by(year, state, month = month(as.Date(poll_date))) |>
  summarize(poll_support = round(weighted.mean(poll_support, weeks_left, na.rm = TRUE), 3), .groups = "drop") |>
  pivot_wider(names_from = month, 
              values_from = poll_support, 
              names_prefix = "month_") |>
  select(year,state, month_9, month_10) |>
  rename(September_Results = month_9, 
         October_Results = month_10) |>
  left_join(kaitlyn_dataset, by = c("state", "year")) |>
  mutate(across(everything(), ~ replace_na(., 0))) 

```

```{r, echo=FALSE, warning=FALSE, message=FALSE, include=FALSE}
prediction_data

```

```{r, echo=FALSE, warning=FALSE, message=FALSE, include=FALSE}
pred44 <- predict(reg44, prediction_data, interval = "prediction")

table44 <- bind_cols(prediction_data,
  as.data.frame(pred44) |> rename_with(~paste0("D_", .))) |>
  mutate(D_fit = round(D_fit, 2), 
         D_lwr = round(D_lwr, 2), 
         D_upr = round(D_upr, 2), 
         winner = if_else(D_fit>50, "Harris", "Trump")) |>
  select(-September_Results,
         -October_Results, 
         -year, ) |>
  filter(state %in% c("Arizona", "Florida", "Georgia", "Michigan", "Minnesota", "Nevada", "New Hampshire", "New Mexico", "North Carolina", "Pennsylvania", "Texas", "Virginia", "Wisconsin"))
```




```{r, echo=FALSE, warning=FALSE, message=FALSE, include=FALSE}

map44 <- expert |>
  mutate(winner = if_else(cook_text_rating == "Solid Democrat", 
                          "Harris", 
                          if_else(cook_text_rating == "Solid Republican", 
                                  "Trump", 
                                  NA_character_))) |>
  slice(-c(52:56)) |>
  select(state, winner) |>
  left_join(table44 |> select(state, winner88 = winner), by = "state") |>
  mutate(winner = coalesce(winner, winner88)) |>
  select(state, winner) |>
  mutate(winner = if_else(state == "Maine", "Harris", winner))
```

```{r, echo=FALSE, warning=FALSE, message=FALSE}
#written in consultation with Kaitlyn Vu
map44 |> 
  left_join(states, by = "state") |>
  ggplot(aes(long, lat, group = group)) + 
  geom_polygon(aes(fill = winner), color = "white", size = 0.4) +  
  scale_fill_manual(values = c(
    "Harris" = "dodgerblue3",
    "Trump" = "firebrick3"    
  )) +
  labs(
    title = "Electoral Projection",
    fill = "Predicted Winner" 
  ) +
  theme_void() +
  theme(
    plot.title = element_text(
      hjust = 0.5, face = "bold", size = 16, color = "black", vjust = 2
    ), 
    legend.title = element_text(size = 12, face = "bold"),
    legend.text = element_text(size = 11),
    legend.key.size = unit(0.8, "cm"),
    legend.position = "bottom",
    plot.margin = margin(20, 20, 20, 20)
  )

```




```{r, echo=FALSE, warning=FALSE, message=FALSE}
electoral_table <- map44 |>
  left_join(d_ec, by = "state") |>
  select(state, winner, electors) |>
  mutate(winner = case_when(
    winner == "Harris" ~ "Kamala Harris",
    winner == "Trump" ~ "Donald Trump",
    TRUE ~ winner
  )) |>
  group_by(winner) |>
  summarize(count = sum(electors)) |>
  arrange(desc(count))


electoral_table |>
  ggplot(aes(x = reorder(winner, -count), y = count, fill = winner)) +
  geom_bar(stat = "identity", width = 0.6, color = "black") +
    geom_hline(yintercept = 270, color = "red", linetype = "solid", size = 1) + 
  scale_fill_manual(values = c(
    "Kamala Harris" = "dodgerblue3",
    "Donald Trump" = "firebrick3"
  )) +
  labs(
    title = "Electoral Vote Distribution",
    x = "Candidate",
    y = "Total Electoral Votes",
    fill = "Candidate"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 14),
    axis.title.x = element_text(size = 12, face = "bold"),
    axis.title.y = element_text(size = 12, face = "bold"),
    axis.text.x = element_text(size = 11),
    axis.text.y = element_text(size = 11),
    legend.position = "none"
  )
```



