---
title: Blog Post 6
author: Package Build
date: '2024-10-12'
slug: blog-post-7
categories: []
tags: []
---
Victor Bowker
Week 6


# Thanks for coming back

Welcome back! This is week 6, and I approve this message.
(get it?)

This week is all about campaign ads!!! Interestingly, my mention of campaign ads may cause groans from some, but others perhaps a mere nod; do you wonder why? Let me tell you!

Campaigns spend incredible amounts of money (NYT says 500 million will be spent between early September and Election Day) but this money is not allocated evenly across the country.[1] In the following graphs, I will first explain how campaigns allocate ad spending across states. Next, I will show you how the tone of these ads can differ. Then I will wrap up with some of my _favorite_ campaign ads, then finally an update to my presidential projection!

Excited? Great, me too, so lets go!


```{r, echo=FALSE, warning = FALSE, message = FALSE, include=FALSE}
library(car)
library(caret)
library(cowplot)
library(curl)
library(CVXR)
library(foreign)
library(geofacet)
library(glmnet)
library(haven)
library(janitor)
library(kableExtra)
library(maps)
library(mlr3)
library(randomForest)
library(ranger)
library(RColorBrewer)
library(rstan)
library(scales)
library(sf)
library(shinystan)
library(tidyverse)
library(viridis)
```

```{r, echo=FALSE, warning = FALSE, message = FALSE, include=FALSE}
# Read popular vote datasets. 
d_popvote <- read_csv("popvote_1948_2020.csv")
d_state_popvote <- read_csv("state_popvote_1948_2020.csv")
```
```{r, echo=FALSE, warning = FALSE, message = FALSE, include=FALSE}
# Read elector distribution dataset. 
d_ec <- read_csv("corrected_ec_1948_2024.csv")
```
```{r, echo=FALSE, warning = FALSE, message = FALSE, include=FALSE}
# Read ads datasets. 
ad_campaigns <- read_csv("ad_campaigns_2000-2012.csv")
ad_creative <- read_csv("ad_creative_2000-2012.csv")
ads_2020 <- read_csv("ads_2020.csv")
facebook_ads_2020 <- read_csv("facebook_ads_2020.csv")
facebook_ads_biden_2020 <- read_csv("facebook_ads_biden_2020.csv")
campaign_spending <- read_csv("FEC_contributions_by_state_2008_2024.csv")
```
```{r, echo=FALSE, warning = FALSE, message = FALSE, include=FALSE}
# Read polling data. 
d_polls <- read_csv("national_polls_1968-2024.csv")
d_state_polls <- read_csv("state_polls_1968-2024.csv")
```
```{r, echo=FALSE, warning = FALSE, message = FALSE, include=FALSE}
# Read turnout data. 
d_turnout <- read_csv("state_turnout_1980_2022.csv")
```

# Ad Tone

Its safe to say that if you are interested enough in the election to read this blog then you probably have seen a campaign ad or two. 

## Embedding a YouTube Video

<iframe width="560" height="315" 
        src="https://www.youtube.com/watch?v=MOEMX6_A8MM" 
        title="YouTube video player" 
        frameborder="0" 
        allow="accelerometer; autoplay; clipboard-write; encrypted-media; gyroscope; picture-in-picture" 
        allowfullscreen>
</iframe>



```{r, echo=FALSE, warning = FALSE, message = FALSE}
# Tone and Political Ads. 
ad_campaigns |>
  left_join(ad_creative) |>
  group_by(cycle, party) |> mutate(tot_n=n()) |> ungroup() |>
  group_by(cycle, party, ad_tone) |> summarise(pct=n()*100/first(tot_n)) |>
  filter(!is.na(ad_tone)) |>
  ggplot(aes(x = cycle, y = pct, fill = ad_tone, group = party)) +
  geom_bar(stat = "identity") +
  scale_x_continuous(breaks = seq(2000, 2012, 4)) +
  ggtitle("Tone of Campaign Ads") +
  scale_fill_manual(values = c("darkblue","blue","steelblue","lightblue","white"), name = "Ad Tone") +
  xlab("Election Year") + ylab("% of Ad Total") +
  facet_wrap(~ party) + theme_minimal() +
    facet_wrap(~ party, labeller = as_labeller(toupper)) +  #found with online research
  theme(axis.title = element_text(size=15),
        axis.text = element_text(size=10),
        strip.text.x = element_text(size = 15)) +
  theme(axis.title = element_text(size = 18, face = "bold"),
        plot.title = element_text(size = 20, face = "bold", hjust = 0.5),
        plot.subtitle = element_text(size = 14, hjust = 0.5),
        legend.position = "top",
        legend.title = element_text(size = 12, face = "bold"),
        legend.text = element_text(size = 10),
        strip.text.x = element_text(size = 15, face = "bold"),
        panel.spacing = unit(1, "lines"))
```
```{r, echo=FALSE, warning = FALSE, message = FALSE, include=FALSE}
## The Purpose of Political Ads
ad_campaigns |>
  left_join(ad_creative) |>
  group_by(cycle, party) |> mutate(tot_n=n()) |> ungroup() |>
  group_by(cycle, party, ad_purpose) |> summarise(pct=n()*100/first(tot_n)) |>
  filter(!is.na(ad_purpose)) |>
  bind_rows( ##2016 raw data not public yet! This was entered manually
    data.frame(cycle = 2016, ad_purpose = "personal", party = "democrat", pct = 67),
    data.frame(cycle = 2016, ad_purpose = "policy", party = "democrat", pct = 12),
    data.frame(cycle = 2016, ad_purpose = "both", party = "democrat", pct = 21),
    data.frame(cycle = 2016, ad_purpose = "personal", party = "republican", pct = 11),
    data.frame(cycle = 2016, ad_purpose = "policy", party = "republican", pct = 71),
    data.frame(cycle = 2016, ad_purpose = "both", party = "republican", pct = 18)
  ) |>
  ggplot(aes(x = cycle, y = pct, fill = ad_purpose, group = party)) +
  geom_bar(stat = "identity") +
  scale_x_continuous(breaks = seq(2000, 2016, 4)) +
  # ggtitle("Campaign Ads Aired By Purpose") +
  scale_fill_manual(values = c("grey","red","darkgreen","black","white"), name = "purpose") +
  xlab("") + ylab("%") +
  facet_wrap(~ party) + theme_minimal() +
  theme(axis.title = element_text(size=20),
        axis.text = element_text(size=15),
        strip.text.x = element_text(size = 20))
```
```{r, echo=FALSE, warning = FALSE, message = FALSE, include=FALSE}
## The Elections and Their Issues
top_issues <- ad_campaigns |> 
  left_join(ad_creative) |>
  filter(!grepl("None|Other", ad_issue)) |>
  group_by(cycle, ad_issue) |> summarise(n=n()) |> top_n(5, n)
```
```{r, echo=FALSE, warning = FALSE, message = FALSE, include=FALSE}
### making each plot in a grid to have its own x-axis (issue name)
### is tricky with `facet_wrap`, so we use this package `cowplot`
### which allows us to take a list of separate plots and grid them together
plist <- lapply(c(2000,2004,2008,2012), function(c) {
  top_issues |> filter(cycle == c) |> 
    ggplot(aes(x = reorder(ad_issue, n), y = n)) +
    geom_bar(stat = "identity") + coord_flip() + theme_bw() +
    xlab("") + ylab("number ads aired") + ggtitle(paste("Top 5 Ad\nIssues in",c))
  
})
cowplot::plot_grid(plotlist = plist, nrow = 2, ncol = 2, align = "hv")
```
```{r, echo=FALSE, warning = FALSE, message = FALSE, include=FALSE}
## Campaign Ads Aired By Issue and Party: 2000
party_issues2000 <- ad_campaigns |>
  filter(cycle == 2000) |>
  left_join(ad_creative) |>
  filter(ad_issue != "None") |>
  ## this `group_by` is to get our denominator
  group_by(ad_issue) |> mutate(tot_n=n()) |> ungroup() |>
  ## this one is get numerator and calculate % by party
  group_by(ad_issue, party) |> summarise(p_n=n()*100/first(tot_n)) |> ungroup() |>
  ## finally, this one so we can sort the issue names
  ## by D% of issue ad-share instead of alphabetically
  group_by(ad_issue) |> mutate(Dp_n = ifelse(first(party) == "democrat", first(p_n), 0))

ggplot(party_issues2000, aes(x = reorder(ad_issue, Dp_n), y = p_n, fill = party)) + 
  geom_bar(stat = "identity") +
  scale_fill_manual(values = c("blue", "red")) +
  ylab("% of ads on topic from each party") + xlab("issue") + 
  # ggtitle("Campaign Ads Aired by Topic in 2000") +
  coord_flip() + 
  theme_bw()
```
```{r, echo=FALSE, warning = FALSE, message = FALSE, include=FALSE}
## Campaign Ads Aired By Issue and Party: 2012
party_issues2012 <- ad_campaigns |>
  filter(cycle == 2012) |>
  left_join(ad_creative) |>
  filter(ad_issue != "None") |>
  group_by(cycle, ad_issue) |> mutate(tot_n=n()) |> ungroup() |>
  group_by(cycle, ad_issue, party) |> summarise(p_n=n()*100/first(tot_n)) |> ungroup() |>
  group_by(cycle, ad_issue) |> mutate(Dp_n = ifelse(first(party) == "democrat", first(p_n), 0))

ggplot(party_issues2012, aes(x = reorder(ad_issue, Dp_n), y = p_n, fill = party)) + 
  geom_bar(stat = "identity") +
  scale_fill_manual(values = c("blue", "red")) +
  ylab("% of ads on topic from each party") + xlab("issue") +
  # ggtitle("Campaign Ads Aired by Topic in 2012") +
  coord_flip() + 
  theme_bw()
```
```{r, echo=FALSE, warning = FALSE, message = FALSE, include=FALSE}
## When to Buy Ads? 
ad_campaigns |>
  mutate(year = as.numeric(substr(air_date, 1, 4))) |>
  mutate(month = as.numeric(substr(air_date, 6, 7))) |>
  filter(year %in% c(2000, 2004, 2008, 2012), month > 7) |>
  group_by(cycle, air_date, party) |>
  summarise(total_cost = sum(total_cost)) |>
  ggplot(aes(x=air_date, y=total_cost, color=party)) +
  # scale_x_date(date_labels = "%b, %Y") +
  scale_y_continuous(labels = dollar_format()) +
  scale_color_manual(values = c("blue","red"), name = "") +
  geom_line() + geom_point(size=0.5) +
  facet_wrap(cycle ~ ., scales="free") +
  xlab("") + ylab("ad spend") +
  theme_bw() +
  theme(axis.title = element_text(size=20),
        axis.text = element_text(size=11),
        strip.text.x = element_text(size = 20))
```
```{r, echo=FALSE, warning = FALSE, message = FALSE, include=FALSE}
## Tone in Political Ads
ad_campaigns |>
  left_join(ad_creative) |>
  filter(ad_tone %in% c("attack", "promote")) |>
  mutate(year = as.numeric(substr(air_date, 1, 4))) |>
  mutate(month = as.numeric(substr(air_date, 6, 7))) |>
  filter(year %in% c(2000, 2004, 2008, 2012), month > 7) |>
  group_by(cycle, air_date, ad_tone) |>
  summarise(total_cost = sum(n_stations)) |>
  group_by(cycle, air_date) |>
  mutate(total_cost = total_cost/sum(total_cost)) |>
  ungroup() |>
  ggplot(aes(x=air_date, y=total_cost, fill=ad_tone, color=ad_tone)) +
  # scale_x_date(date_labels = "%b") +
  scale_fill_manual(values = c("purple","green"), name = "ad tone") +
  scale_color_manual(values = c("purple","green"), name = "ad tone") +
  geom_bar(stat = "identity") +
  facet_wrap(cycle ~ ., scales="free") +
  xlab("") + ylab("% of ads bought on day") +
  theme_bw() +
  theme(axis.title = element_text(size=20),
        axis.text = element_text(size=10),
        strip.text.x = element_text(size = 20))
```
```{r, echo=FALSE, warning = FALSE, message = FALSE, include=FALSE}
## The State-level Air War in 2008 (Obama vs. McCain)
ad_campaigns |>
  mutate(year = as.numeric(substr(air_date, 1, 4))) |>
  mutate(month = as.numeric(substr(air_date, 6, 7))) |>
  mutate(state = state.name[match(state, state.abb)]) |>
  filter(cycle == 2008) |>
  left_join(d_state_popvote |> filter(year == 2008) |> select(-year), by="state") |>
  mutate(winner=ifelse(D_pv2p > R_pv2p, "democrat", "republican")) |>
  group_by(cycle, state, air_date, party, winner) |>
  summarise(total_cost = sum(total_cost)) |>
  filter(!is.na(state)) |>
  # ggplot(aes(x=air_date, y=log(total_cost+1), color=party)) +
  ggplot(aes(x=party, y=total_cost, fill=party)) +
  geom_bar(stat="identity") +
  geom_rect(aes(fill=winner), xmin=-Inf, xmax=Inf, ymin=46.3*10^6, ymax=52*10^6) +
  facet_geo(~ state, scales="free_x") +
  scale_fill_manual(values = c("blue", "red")) +
  scale_y_continuous(labels = unit_format(unit = "M", scale = 1e-6)) +
  xlab("") + ylab("ad spend") +
  theme_bw() +
  theme(axis.title.x=element_blank(), 
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank())

```
```{r, echo=FALSE, warning = FALSE, message = FALSE, include=FALSE}
# Visualizing Facebook ads and Biden Facebook ads in 2020. 
d_facebook <- facebook_ads_2020 |> 
  rename(date = from_date, new_ads = num_of_new_ads) |> 
  group_by(date) |> 
  summarize(new_spend = sum(new_spend, na.rm = T),
            new_ads = sum(new_ads, na.rm = T))

d_facebook |> 
  ggplot(aes(x = date, y = new_ads)) + 
  geom_line() +
  geom_smooth(method = "lm", se = TRUE) +
  labs(x = "Date", 
       y = "New Facebook Ads") +
  theme_minimal()

d_facebook |> 
  ggplot(aes(x = date, y = new_spend)) +
  geom_line() +
  scale_y_continuous(labels = dollar_format()) +
  geom_smooth(method = "lm", se = TRUE) + 
  labs(x = "Date", 
       y = "New Facebook Ad Spending") +
  theme_minimal()
  
d_facebook_biden <- facebook_ads_biden_2020 |> 
  rename(date = from_date, new_ads = num_of_new_ads) |> 
  group_by(date) |> 
  summarize(new_spend = sum(new_spend, na.rm = T),
            new_ads = sum(new_ads, na.rm = T))

d_facebook_biden |>
  ggplot(aes(x = date, y = new_ads)) + 
  geom_line() +
  geom_smooth(method = "lm", se = TRUE) +
  labs(x = "Date", 
       y = "New Facebook Ads (Biden Only)") +
  theme_minimal()

d_facebook_biden |>
  ggplot(aes(x = date, y = new_spend)) +
  geom_line() +
  scale_y_continuous(labels = dollar_format()) +
  geom_smooth(method = "lm", se = TRUE) + 
  labs(x = "Date", 
       y = "New Facebook Ad Spending (Biden Only)")
  theme_minimal()
```


Citations
[1] - https://www.nytimes.com/2024/09/17/us/elections/presidential-campaign-advertising-spending.html



