---
title: Blog Post 8
author: Victor Bowker
date: '2024-10-26'
slug: blog-post-8
categories: []
tags: []
---

# Welcome Back

Are you getting excited? This election is _SO SOON_, and I do not know how to feel. I am certainly a nerd for politics, so I am excited to watch the results roll in, but there are undoubtedly great expectations and implications for who will win. 

If you are wondering how many days until that fateful Tuesday, check here! (This was written on Saturday, October 26th!)

```{r, echo=FALSE, warning=FALSE, message=FALSE}
# How many days until the election? 
election.day <- as.Date("2024-11-05")
current.date <- Sys.Date()

election.day-current.date
# 20 days!!!!!!
```

```{r, include=FALSE, echo=FALSE, warning=FALSE, message=FALSE}
options(repos = c(CRAN = "https://cloud.r-project.org"))

```

# Shocks

This week, in 1347, we discussed how shocks can impact an election outcome. Perhaps you are wondering, _what is a shock_?

Well, let me tell you! First, I will discuss apolitical shocks--things that occur outside of the partisan realm. This can include natural disasters (Healy and Malhotra 2010), as well as shark attacks (Achen and Bartels 2017) or even the outcome of a big sports game (Healy, Mo, Malhotra 2010)

  When I think of events in this realm, some of the standouts include recent hurricanes in the South, namely Milton and Helene. It feels reasonable to say that these events could greatly impact the public opinion of local officials. A recent article from the AP claims that North Carolina has summed up the damage caused from Hurricane Helene, and the total cost is nearly $53 billion (1). That is a staggering number by any measure, and the state is in great peril. How officials, like Governor Roy Cooper, and President Joe Biden, respond to this crisis could potentially implicate them in future elections (though of course we know President Biden is not seeking re-election). This apolitical event is causing the people of North Carolina to suffer incredibly, and it serves as a test of the leadership's reactionary abilities. 

  Another more fun example of an apolitical shock is a shark attack. An example, introduced by Achen and Bartels in their 2012 piece, shows that after a "dramatic series of shark attacks" on the shores of New Jersey in 1916, voters "significantly punished" President Woodrow Wilson in his re-election race. That sounds a little _far fetched_, right? (2) The Princeton Professors argued that voters are truly less evaluative of actual foreign or domestic issues than previously believed. They say that instead of true political or governmental issues, local or suprising shocks, like a shark attack, are more likely to influence the results of the election.

  Because someone had to do it, Fowler and Hall later reevaluated the above theories. They found some _less comical_ results, showing that there is in fact no "systematic evidence that shark attacks affect elections" - sad, right? (3) Fowler and Hall came up with a few reasons for the viral shark attack paper to have resulted in those conclusions, including the potential for an ineffectively controlled experiment that did not consider a variety of other factors, or even worse, that potentially Achen and Bartels cherry picked data to find something fitting their newsworthy conclusion. 
  
  That is a hefty accusation, and I am in no place to take a side, but let me leave you with this. Last week, in Florida, a young man was attacked by a shark when surfing. (4) If I was feeling particularly Achen-ish, I might declare this to be the reason the incumbent Democrats won't win Florida this time--but I bet you could come up with a few other reasons why Harris won't receive Florida's 30 Electoral Votes. But, in any case, now I have published my declaration that this shark attack will flip Florida (but lets see if I come back on November 6th and delete this post)

  I mentioned apolitical events, but what about purely political ones? In the past week, Former President Trump's Former Chief of Staff, John Kelly, claimed that if elected, Trump would "meet the definition of a fascist" and historically has glorified the actions of Adolf Hitler. (5) Now, in the past few days, 13 former Trump aides ended their potential for a job in another Trump admin by releasing a statement echoing the worries of Kelly. (6) These statements have been discussed widely in the past week, but one thing working against the validity of the claims is at least a few of the signers are outspoken Harris advocates, paving the way for Trump to write this off as a political move. Examples of historical political events include in Comey Letter in the days leading up to the 2016 Election, where FBI Director James Comey notified Congress that the bureau had launched an investigation into then candidate for President, Hillary Clinton. Just da4ys before the election, 538 analysts have attributed a "3 or 4 percentage point shift towards Donald Trump, swinging Michigan, Pennsylvania, Wisconsin, and Florida." (7)
  
  With that, *do you think these shock events will sway the election?* I, for one, am not too sure.

# Projection Time!

We are less than 10 days away from the 2024 Presidential Election. I have been thinking about this race since the time I have started to think, or since 2020, whichever came second :). 
```{r, include=FALSE, echo=FALSE, warning=FALSE, message=FALSE}
install.packages("patchwork")
library(patchwork)
library(broom)
library(geofacet)
library(ggpubr)
library(ggthemes)
library(haven)
library(DT)
library(maps)
library(mgcv)
library(mgcViz)
library(RColorBrewer)
library(scales)
library(sf)
library(spData)
library(stargazer)
library(tidygeocoder)
library(tidyverse)
library(tigris)
library(tmap)
library(tmaptools)
library(viridis)
library(tidyverse)
library(gt)
install.packages("readxl")
library(readxl)
```

```{r, include=FALSE, echo=FALSE, warning=FALSE, message=FALSE}
# Read popular vote datasets. 
d_popvote <- read_csv("popvote_1948_2020.csv")
d_state_popvote <- read_csv("state_popvote_1948_2020.csv")
states <- map_data("state") |>
  rename(state = region)
states$state <- str_to_title(states$state)
d_state_popvote[d_state_popvote$state == "District of Columbia",]$state <- "District Of Columbia"
```

```{r, include=FALSE, echo=FALSE, warning=FALSE, message=FALSE}
# Read elector distribution dataset. 
d_ec <- read_csv("corrected_ec_1948_2024.csv")
d_ec <- d_ec |>
  filter(year == 2024)
library(readxl)
expert <- read_excel("/Users/victorbowker/Fall 2024/election-blog/content/post/2024-10-26-blog-post-8/expert_2024-2.xlsx")

```

```{r, include=FALSE, echo=FALSE, warning=FALSE, message=FALSE}
# Read polling data. 
d_polls <- read_csv("national_polls_1968-2024.csv")
d_state_polls <- read_csv("state_polls_1968-2024.csv")
```

```{r, include=FALSE, echo=FALSE, warning=FALSE, message=FALSE}
# Read turnout data. 
d_turnout <- read_csv("state_turnout_1980_2022.csv")
```

```{r, include=FALSE, echo=FALSE, warning=FALSE, message=FALSE}
# Read county turnout. 
d_county_turnout <- read_csv("county_turnout.csv")
```

```{r, include=FALSE, echo=FALSE, warning=FALSE, message=FALSE}
# Read state-level demographics.
d_state_demog <- read_csv("demographics.csv")
```

```{r, include=FALSE, echo=FALSE, warning=FALSE, message=FALSE}
# Read county demographics. 
d_county_demog <- read_csv("county_demographics.csv")
```

```{r, include=FALSE, echo=FALSE, warning=FALSE, message=FALSE}
# Read campaign events datasets. 
d_campaign_events <- read_csv("campaigns_2016_2024.csv")[,-1]
```

```{r, echo=FALSE, warning=FALSE, message=FALSE, include=FALSE}

state_abb_xwalk <- d_state_popvote |>
  mutate(state_abb = state.abb[match(d_state_popvote$state, state.name)]) |> 
  select(state, state_abb) |> 
  distinct() 
state_abb_xwalk[51,]$state <- "District Of Columbia"
state_abb_xwalk[51,]$state_abb <- "DC"
```

```{r, echo=FALSE, warning=FALSE, message=FALSE, include=FALSE}
d_campaign_events <-
  d_campaign_events |>
  rename("state_abb" = "state")
```

```{r, echo=FALSE, warning=FALSE, message=FALSE, include=FALSE}
d_campaign_events$party[d_campaign_events$candidate %in% c("Trump / Pence", "Trump", "Pence", "Trump/Pence", "Vance")] <- "REP"
d_campaign_events$party[d_campaign_events$candidate %in% c("Biden / Harris", "Biden", "Harris", "Biden/Harris", "Walz", "Kaine", "Clinton", "Clinton / Kaine")] <- "DEM"
```


```{r, echo=FALSE, warning=FALSE, message=FALSE, include=FALSE}
kaitlyn_dataset <- 
  d_campaign_events |> 
  filter(party == "DEM") |>
  left_join(state_abb_xwalk, by = "state_abb") |> 
  left_join(d_state_popvote, by = c("state", "year")) |>
  group_by(state, year) |>
  summarize(fun_campaign_events = n()) 

kaitlyn_dataset
```

```{r, echo=FALSE, warning=FALSE, message=FALSE, include=FALSE}
library(dplyr)
library(tidyr)
library(lubridate)

model88 <- d_state_polls |>
  select(state, year, weeks_left, poll_date, poll_support, party) |>
  filter(month(as.Date(poll_date)) %in% c(9, 10), party == "DEM", year >= 2016) |>
  filter(year < 2024) |>
  group_by(year, state, month = month(as.Date(poll_date))) |>
  summarize(poll_support = round(weighted.mean(poll_support, weeks_left, na.rm = TRUE), 3), .groups = "drop") |>
  pivot_wider(names_from = month, 
              values_from = poll_support, 
              names_prefix = "month_") |>
  left_join(d_state_popvote, by = c("year", "state")) |>
  select(year,state, month_9, month_10, D_pv2p) |>
  rename(September_Results = month_9, 
         October_Results = month_10) |>
  left_join(d_popvote |>
  select(year, party, incumbent_party) |>
  filter(party == "DEM") |>
  mutate(incumbent_party = if_else(incumbent_party == "TRUE", 1, 0)) |>
  select(-party), by = "year")



```

```{r, echo=FALSE, warning=FALSE, include=FALSE, message=FALSE}
reg88 <- lm(D_pv2p ~ September_Results + October_Results, 
                data = subset(model88, year <= 2020))
```


```{r, echo=FALSE, warning=FALSE, message=FALSE, include=FALSE}
tidy_model88 <- tidy(reg88)
```



As with most of my past posts, and with the general understanding of voters making decisions based on recent events and opinions, October is must more significant than September. See below!


```{r, echo=FALSE, warning=FALSE, message=FALSE}
datatable(tidy_model88, 
          options = list(pageLength = 4), 
          caption = "Updated Regression Results") 
```


```{r, echo=FALSE, warning=FALSE, message=FALSE, include=FALSE}

prediction_data88 <- d_state_polls |>
  select(year, state, weeks_left, poll_date, poll_support, party) |>
  filter(month(poll_date) %in% c(9, 10),
         party == "DEM", 
         year == 2024) |>
  group_by(year, state, month = month(poll_date)) |>
  summarize(poll_support = round(weighted.mean(poll_support, weeks_left, na.rm = TRUE), 2), .groups = "drop") |>
  pivot_wider(names_from = month, 
              values_from = poll_support, 
              names_prefix = "month_") |>
  drop_na() |>
  rename(September_Results = month_9, 
         October_Results = month_10)

```

```{r, echo=FALSE, warning=FALSE, message=FALSE, include=FALSE}
prediction_data88

```

```{r, echo=FALSE, warning=FALSE, message=FALSE, include=FALSE}
pred88 <- predict(reg88, prediction_data88, interval = "prediction")

table88 <- bind_cols(prediction_data88,
  as.data.frame(pred88) |> rename_with(~paste0("D_", .))) |>
  mutate(D_fit = round(D_fit, 2), 
         D_lwr = round(D_lwr, 2), 
         D_upr = round(D_upr, 2), 
         winner = if_else(D_fit>50, "Harris", "Trump")) |>
  select(-September_Results,
         -October_Results, 
         -year, ) |>
  filter(state %in% c("Arizona", "Florida", "Georgia", "Michigan", "Minnesota", "Nevada", "New Hampshire", "New Mexico", "North Carolina", "Pennsylvania", "Texas", "Virginia", "Wisconsin"))
```

```{r, echo=FALSE, warning=FALSE, message=FALSE}
knitr::kable(table88, 
             col.names = c("State", "Prediction", "Lower Bound", "Upper Bound", "Winner")) |>
  kableExtra::kable_styling(full_width = FALSE) |>
  kableExtra::row_spec(row = which(table88$winner == "Harris"), background = "blue") |>
  kableExtra::row_spec(row = which(table88$winner == "Trump"), background = "red") 
```
```{r, echo=FALSE, warning=FALSE, message=FALSE, include=FALSE}

map88 <- expert |>
  mutate(winner = if_else(cook_text_rating == "Solid Democrat", 
                          "Harris", 
                          if_else(cook_text_rating == "Solid Republican", 
                                  "Trump", 
                                  NA_character_))) |>
  slice(-c(52:56)) |>
  select(state, winner) |>
  left_join(table88 |> select(state, winner88 = winner), by = "state") |>
  mutate(winner = coalesce(winner, winner88)) |>
  select(state, winner) |>
  mutate(winner = if_else(state == "Maine", "Harris", winner))
```



# Map!

Thanks to my dear friend Kaitlyn Vu's incredible Excel abilities, I have been able to now make a map! She combined expert predictions and historical data to fill in a few blanks-meaning I am pretty confident that Massachusetts will remain blue, so I won't run a regression on that. This sheet has all the solid states already imported!

Below you will see a map of the election outcome per my latest prediction. First big thing to notice is the most important state(at least in my opinion), Pennsylvania, goes to Harris. She also seized Michigan, Wisconsin, North Carolina, Georgia, Arizona, and basically every other toss up. So, you probably already know, but keep scrolling to see who I have forecasted to win this race.

```{r, echo=FALSE, warning=FALSE, message=FALSE}
#written in consultation with Kaitlyn Vu
map88 |> 
  left_join(states, by = "state") |>
  ggplot(aes(long, lat, group = group)) + 
  geom_polygon(aes(fill = winner), color = "white", size = 0.4) +  
  scale_fill_manual(values = c(
    "Harris" = "dodgerblue3",
    "Trump" = "firebrick3"    
  )) +
  labs(
    title = "Electoral Projection",
    fill = "Predicted Winner" 
  ) +
  theme_void() +
  theme(
    plot.title = element_text(
      hjust = 0.5, face = "bold", size = 16, color = "black", vjust = 2
    ), 
    legend.title = element_text(size = 12, face = "bold"),
    legend.text = element_text(size = 11),
    legend.key.size = unit(0.8, "cm"),
    legend.position = "bottom",
    plot.margin = margin(20, 20, 20, 20)
  )

```

# Winner: Harris

As of now, I have Harris with 319 Electoral Votes, compared to Trump's 219. VP Harris easily surpasses the 270 threshold in this situation and takes the White House!

On a less analytical note, I have been watching forecasts from many more professional sources who are starting to give Trump the upper hand. It is obvious this will be a close race, but perhaps closer than I expected.
```{r, echo=FALSE, warning=FALSE, message=FALSE}
electoral_table <- map88 |>
  left_join(d_ec, by = "state") |>
  select(state, winner, electors) |>
  mutate(winner = case_when(
    winner == "Harris" ~ "Kamala Harris",
    winner == "Trump" ~ "Donald Trump",
    TRUE ~ winner
  )) |>
  group_by(winner) |>
  summarize(count = sum(electors)) |>
  arrange(desc(count))


electoral_table |>
  ggplot(aes(x = reorder(winner, -count), y = count, fill = winner)) +
  geom_bar(stat = "identity", width = 0.6, color = "black") +
    geom_hline(yintercept = 270, color = "red", linetype = "solid", size = 1) + 
  scale_fill_manual(values = c(
    "Kamala Harris" = "dodgerblue3",
    "Donald Trump" = "firebrick3"
  )) +
  labs(
    title = "Electoral Vote Distribution",
    x = "Candidate",
    y = "Total Electoral Votes",
    fill = "Candidate"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 14),
    axis.title.x = element_text(size = 12, face = "bold"),
    axis.title.y = element_text(size = 12, face = "bold"),
    axis.text.x = element_text(size = 11),
    axis.text.y = element_text(size = 11),
    legend.position = "none"
  )
```
 
# What If They Tie?
 
 I mentioned the race is close-but what if it is so close that they tie? According to CNN, "If Harris wins Wisconsin, Michigan, Arizona and Nevada and a single electoral vote in Nebraska, all of which Joe Biden won in 2020, but she loses Pennsylvania and Georgia, there's a tie, 269-269." (8) 
 If this does in fact happen, the 12th Amendment will kick in and allow the newly elected Congress to vote for the next president. This _probably_ won't happen, but if it did, it would be a great time to be concentrating in Government!
 
 
 Citations:
 
 1 - https://apnews.com/article/north-carolina-hurricane-helene-damage-estimate-1c32acaf158bf17eae815926ea682c0e
 2-  https://www.vanderbilt.edu/csdi/research/CSDI_WP_05-2013.pdf
 3 - https://www.journals.uchicago.edu/doi/abs/10.1086/699244
4 - https://www.foxnews.com/us/florida-teen-surfer-attacked-shark-says-could-have-been-so-much-worse
5 - https://www.nytimes.com/2024/10/22/us/politics/john-kelly-trump-fitness-character.html
6 - https://www.nytimes.com/2024/10/25/us/politics/trump-officials-letter-fascist-john-kelly.html
7 - https://fivethirtyeight.com/features/the-comey-letter-probably-cost-clinton-the-election/
8 - https://www.cnn.com/2024/08/04/politics/tie-presidential-election-what-matters/index.html
9 - Kaitlyn Vu - check her out here -> https://kaitvu.github.io/election-blog/

 
 
 
 
 
 