---
title: Blog Post 2
author: Victor Bowker
date: '2024-09-14'
slug: blog-post-2
categories: []
tags: []
---

## Introduction
It is no suprise, nor a secret, that most Americans vote with their pocketbook in mind. 
According to the Pew Research Center in a poll from September 9th of this year, the _greatest_ concern among both Trump and Harris supporters is the economy. The next greatest issue, health care, is only ranked at highly concerning to 65% of voters, compared to the 81% when asked about the American economy.


```{r,echo=FALSE, include=FALSE, warning = FALSE, message = FALSE}
library(car)
library(tidyverse)
library(readr)

```
```{r,echo=FALSE, include=FALSE, warning = FALSE, message = FALSE}
d_popvote <- read_csv("popvote_1948-2020.csv")
```

```{r,echo=FALSE, include=FALSE, warning = FALSE, message = FALSE}
d_fred <- read_csv("fred_econ.csv")
```

```{r,echo=FALSE, include=FALSE, warning = FALSE, message = FALSE}
d_bea <- read_csv("bea_econ.csv") |> 
  rename(year = "Year",
         quarter = "Quarter", 
         gdp = "Gross domestic product", 
         gnp = "Gross national product", 
         dpi = "Disposable personal income", 
         consumption = "Personal consumption expenditures", 
         goods = "Goods", 
         durables = "Durable goods", 
         nondurables = "Nondurable goods", 
         services = "Services", 
         pop = "Population (midperiod, thousands)")
```


```{r,echo=FALSE, include=FALSE, warning = FALSE, message = FALSE}
d_inc_econ <- d_popvote |> 
  filter(incumbent_party == TRUE) |> 
  select(year, pv, pv2p, winner) |> 
  left_join(d_fred |> filter(quarter == 2)) |> 
  left_join(d_bea |> filter(quarter == "Q2") |> select(year, dpi))
```

```{r, echo=FALSE, warning = FALSE, message = FALSE}
d_inc_econ |> 
  ggplot(aes(x = GDP_growth_quarterly, y = pv2p, label = year)) + 
  geom_text() + 
  geom_hline(yintercept = 50, lty = 2) + 
  geom_vline(xintercept = 0.01, lty = 2) +
  labs(x = "Second Quarter GDP Growth (%)", 
       y = "Incumbent Party's National Popular Vote Share") + 
  theme_bw()
```

```{r,echo=FALSE, include=FALSE, warning = FALSE, message = FALSE}
d_inc_econ_2 <- d_inc_econ |>
  filter(year != 2020)
```

```{r, echo=FALSE, warning = FALSE, message = FALSE}
d_inc_econ_2 |> 
  ggplot(aes(x = GDP_growth_quarterly, y = pv2p, label = year)) + 
  geom_text() + 
  geom_hline(yintercept = 50, lty = 2) + 
  geom_vline(xintercept = 0.01, lty = 2) + 
  labs(x = "Second Quarter GDP Growth (%)", 
       y = "Incumbent Party's National Popular Vote Share") + 
  theme_bw()
```

```{r,echo=FALSE, include=FALSE, warning = FALSE, message = FALSE}
cor(d_inc_econ$GDP_growth_quarterly, 
    d_inc_econ$pv2p)
cor(d_inc_econ_2$GDP_growth_quarterly, 
    d_inc_econ_2$pv2p)
```
```{r,echo=FALSE, include=FALSE, warning = FALSE, message = FALSE}
reg_econ <- lm(pv2p ~ GDP_growth_quarterly, 
               data = d_inc_econ)
reg_econ |> summary()

reg_econ_2 <- lm(pv2p ~ GDP_growth_quarterly, 
                         data = d_inc_econ_2)
reg_econ_2 |> summary()
```

```{r, echo=FALSE, warning = FALSE, message = FALSE}
d_inc_econ |> 
  ggplot(aes(x = GDP_growth_quarterly, y = pv2p, label = year)) + 
  geom_text() + 
  geom_smooth(method = "lm", formula = y ~ x) +
  geom_hline(yintercept = 50, lty = 2) + 
  geom_vline(xintercept = 0.01, lty = 2) + 
  labs(x = "Second Quarter GDP Growth (%)", 
       y = "Incumbent Party's National Popular Vote Share", 
       title = "Y = 51.25 + 0.274 * X") + 
  theme_bw() + 
  theme(plot.title = element_text(size = 18))
```

```{r, echo=FALSE, warning = FALSE, message = FALSE}
d_inc_econ_2 |> 
  ggplot(aes(x = GDP_growth_quarterly, y = pv2p, label = year)) + 
  geom_text() + 
  geom_smooth(method = "lm", formula = y ~ x) +
  geom_hline(yintercept = 50, lty = 2) + 
  geom_vline(xintercept = 0.01, lty = 2) + 
  labs(x = "Second Quarter GDP Growth (%)", 
       y = "Incumbent Party's National Popular Vote Share", 
       title = "Y = 49.38 + 0.737 * X") + 
  theme_bw() + 
  theme(plot.title = element_text(size = 18))
```

```{r,echo=FALSE, include=FALSE, warning = FALSE, message = FALSE}
summary(reg_econ_2)$r.squared
summary(reg_econ_2)$adj.r.squared
```

```{r,echo=FALSE, warning = FALSE, message = FALSE}
plot(d_inc_econ$year, 
     d_inc_econ$pv2p, 
     type="l",
     main="True Y (Line), Predicted Y (Dot) for Each Year")
points(d_inc_econ$year, predict(reg_econ_2, d_inc_econ))
```

```{r, echo=FALSE, warning = FALSE, message = FALSE}
plot(reg_econ_2)
```

```{r, echo=FALSE, warning = FALSE, message = FALSE}
hist(reg_econ_2$model$pv2p - reg_econ_2$fitted.values, 
     main = "Histogram of True Y - Predicted Y")
mse <- mean((reg_econ_2$model$pv2p - reg_econ_2$fitted.values)^2)
mse
sqrt(mse)
```
```{r, echo=FALSE, warning = FALSE, message = FALSE}
(out_samp_pred <- predict(reg_econ_2, d_inc_econ[d_inc_econ$year == 2020,]))
(out_samp_truth <- d_inc_econ |> filter(year == 2020) |> select(pv2p))
out_samp_pred - out_samp_truth 
```
```{r, echo=FALSE, warning = FALSE, message = FALSE}
years_out_samp <- sample(d_inc_econ_2$year, 9) 
mod <- lm(pv2p ~ GDP_growth_quarterly, 
          d_inc_econ_2[!(d_inc_econ_2$year %in% years_out_samp),])
out_samp_pred <- predict(mod, d_inc_econ_2[d_inc_econ_2$year %in% years_out_samp,])
out_samp_truth <- d_inc_econ_2$pv2p[d_inc_econ_2$year %in% years_out_samp]
mean(out_samp_pred - out_samp_truth)
```

```{r, echo=FALSE, warning = FALSE, message = FALSE}
out_samp_errors <- sapply(1:1000, function(i) {
  years_out_samp <- sample(d_inc_econ_2$year, 9) 
  mod <- lm(pv2p ~ GDP_growth_quarterly, 
            d_inc_econ_2[!(d_inc_econ_2$year %in% years_out_samp),])
  out_samp_pred <- predict(mod, d_inc_econ_2[d_inc_econ_2$year %in% years_out_samp,])
  out_samp_truth <- d_inc_econ_2$pv2p[d_inc_econ_2$year %in% years_out_samp]
  mean(out_samp_pred - out_samp_truth)
})

mean(abs(out_samp_errors))

hist(out_samp_errors,
     xlab = "",
     main = "Mean Out-of-Sample Residual\n(1000 Runs of Cross-Validation)")
```

```{r, echo=FALSE, warning = FALSE, message = FALSE}
GDP_new <- d_fred |> 
  filter(year == 2024 & quarter == 2) |> 
  select(GDP_growth_quarterly)
```

```{r, echo=FALSE, warning = FALSE, message = FALSE}
predict(reg_econ_2, GDP_new)
```

```{r, echo=FALSE, warning = FALSE, message = FALSE}
predict(reg_econ_2, GDP_new, interval = "prediction")
```



