---
title: Blog Post 2
author: Victor Bowker
date: '2024-09-14'
slug: blog-post-2
categories: []
tags: []
---

## Introduction
It is no surprise, nor a secret, that most Americans vote with their pocketbook in mind. 
According to the Pew Research Center in a poll from September 9th of this year, the _greatest_ concern among both Trump and Harris supporters is the economy. The next greatest issue, health care, is only ranked at highly concerning to 65% of voters, compared to the 81% when asked about the American economy.[1]

Today, I will walk you through analysis of the American economy, how it has historically connected with elections, and how I expect it will influence future elections. 


```{r,echo=FALSE, include=FALSE, warning = FALSE, message = FALSE}
library(car)
library(tidyverse)
library(readr)

```
```{r,echo=FALSE, include=FALSE, warning = FALSE, message = FALSE}
d_popvote <- read_csv("popvote_1948-2020.csv")
```

```{r,echo=FALSE, include=FALSE, warning = FALSE, message = FALSE}
d_fred <- read_csv("fred_econ.csv")
```

```{r,echo=FALSE, include=FALSE, warning = FALSE, message = FALSE}
d_bea <- read_csv("bea_econ.csv") |> 
  rename(year = "Year",
         quarter = "Quarter", 
         gdp = "Gross domestic product", 
         gnp = "Gross national product", 
         dpi = "Disposable personal income", 
         consumption = "Personal consumption expenditures", 
         goods = "Goods", 
         durables = "Durable goods", 
         nondurables = "Nondurable goods", 
         services = "Services", 
         pop = "Population (midperiod, thousands)")
```


```{r,echo=FALSE, include=FALSE, warning = FALSE, message = FALSE}
d_inc_econ <- d_popvote |> 
  filter(incumbent_party == TRUE) |> 
  select(year, pv, pv2p, winner) |> 
  left_join(d_fred |> filter(quarter == 2)) |> 
  left_join(d_bea |> filter(quarter == "Q2") |> select(year, dpi))
```


## Understanding Post Results

  In the graph below, you will see historical trends with the X-axis consisting of GDP growth by quarter, and the Y-axis indicating the incumbent party's national popular vote share. Specifically, the data in the X-axis comes from the second economic quarter (Q2) of an election year. 
  Q2 was selected based on the idea of retrospective voting, where citizens consider past (and generally quite recent) history to decide how to vote now. In 1980, President Reagan embodied the retrospective voter well, asking "are you better off now than you were four years ago?" He reminded voters that incumbent Jimmy Carter was to blame for the past four years of turmoil, and who to celebrate for the past four of accomplishments. Luckily for Reagan, voters weighed Carter's failures higher than his accomplishments, eventually leading to Reagan's November win. You will see in the graph below how our data demonstrates that 1980 result.
  One clear issue is that of 2020, which perhaps adds a flavor of confusion to the seemingly positive correlation. So, what happened in 2020? Growth was slow, as Covid-19 led to some tough times financially. Companies were still shuttered, and it was not easy to both obey social distancing restrictions and run a functioning business. This does, however, make sense when you consider that growth was down, so the incumbent received a low share of the popular vote. Just because the difference in GDP growth is so substantial compared to other years, the general idea holds true! You will see the second graph in fact removes 2020 from the equation to draw a stronger picture.

```{r, echo=FALSE, warning = FALSE, message = FALSE}
d_inc_econ |> 
  ggplot(aes(x = GDP_growth_quarterly, y = pv2p, label = year)) + 
  geom_text() + 
  geom_hline(yintercept = 50, lty = 2) + 
  geom_vline(xintercept = 0.01, lty = 2) +
  labs(x = "Second Quarter GDP Growth (%)", 
       y = "Incumbent Party's National Popular Vote Share",
       title = "Past Trends of GDP and Incumbent Vote Share (1948-2020") + 
  theme_bw()
```

```{r,echo=FALSE, include=FALSE, warning = FALSE, message = FALSE}
d_inc_econ_2 <- d_inc_econ |>
  filter(year != 2020)
```

```{r, echo=FALSE, warning = FALSE, message = FALSE}
d_inc_econ_2 |> 
  ggplot(aes(x = GDP_growth_quarterly, y = pv2p, label = year)) + 
  geom_text() + 
  geom_hline(yintercept = 50, lty = 2) + 
  geom_vline(xintercept = 0.01, lty = 2) + 
  labs(x = "Second Quarter GDP Growth (%)", 
       y = "Incumbent Party's National Popular Vote Share",
       title = "Past Trends of GDP and Incumbent Vote Share (1948-2016") + 
  theme_bw()
```

```{r,echo=FALSE, include=FALSE, warning = FALSE, message = FALSE}
cor(d_inc_econ$GDP_growth_quarterly, 
    d_inc_econ$pv2p)
cor(d_inc_econ_2$GDP_growth_quarterly, 
    d_inc_econ_2$pv2p)
```
```{r,echo=FALSE, include=FALSE, warning = FALSE, message = FALSE}
reg_econ <- lm(pv2p ~ GDP_growth_quarterly, 
               data = d_inc_econ)
reg_econ |> summary()

reg_econ_2 <- lm(pv2p ~ GDP_growth_quarterly, 
                         data = d_inc_econ_2)
reg_econ_2 |> summary()
```

```{r, echo=FALSE, warning = FALSE, message = FALSE}
d_inc_econ |> 
  ggplot(aes(x = GDP_growth_quarterly, y = pv2p, label = year)) + 
  geom_text() + 
  geom_smooth(method = "lm", formula = y ~ x) +
  geom_hline(yintercept = 50, lty = 2) + 
  geom_vline(xintercept = 0.01, lty = 2) + 
  labs(x = "Second Quarter GDP Growth (%)", 
       y = "Incumbent Party's National Popular Vote Share", 
       title = "Y = 51.25 + 0.274 * X") + 
  theme_bw() + 
  theme(plot.title = element_text(size = 18))
```

```{r, echo=FALSE, warning = FALSE, message = FALSE}
d_inc_econ_2 |> 
  ggplot(aes(x = GDP_growth_quarterly, y = pv2p, label = year)) + 
  geom_text() + 
  geom_smooth(method = "lm", formula = y ~ x) +
  geom_hline(yintercept = 50, lty = 2) + 
  geom_vline(xintercept = 0.01, lty = 2) + 
  labs(x = "Second Quarter GDP Growth (%)", 
       y = "Incumbent Party's National Popular Vote Share", 
       title = "Y = 49.38 + 0.737 * X") + 
  theme_bw() + 
  theme(plot.title = element_text(size = 18))
```

```{r,echo=FALSE, include=FALSE, warning = FALSE, message = FALSE}
summary(reg_econ_2)$r.squared
summary(reg_econ_2)$adj.r.squared
```

```{r,echo=FALSE, warning = FALSE, message = FALSE}
plot(d_inc_econ$year, 
     d_inc_econ$pv2p, 
     type="l",
     main="True Y (Line), Predicted Y (Dot) for Each Year")
points(d_inc_econ$year, predict(reg_econ_2, d_inc_econ))
```

```{r, echo=FALSE, warning = FALSE, message = FALSE}
plot(reg_econ_2)
```

```{r, echo=FALSE, warning = FALSE, message = FALSE}
hist(reg_econ_2$model$pv2p - reg_econ_2$fitted.values, 
     main = "Histogram of True Y - Predicted Y")
mse <- mean((reg_econ_2$model$pv2p - reg_econ_2$fitted.values)^2)
mse
sqrt(mse)
```
```{r, echo=FALSE, warning = FALSE, message = FALSE}
(out_samp_pred <- predict(reg_econ_2, d_inc_econ[d_inc_econ$year == 2020,]))
(out_samp_truth <- d_inc_econ |> filter(year == 2020) |> select(pv2p))
out_samp_pred - out_samp_truth 
```
```{r, echo=FALSE, warning = FALSE, message = FALSE}
years_out_samp <- sample(d_inc_econ_2$year, 9) 
mod <- lm(pv2p ~ GDP_growth_quarterly, 
          d_inc_econ_2[!(d_inc_econ_2$year %in% years_out_samp),])
out_samp_pred <- predict(mod, d_inc_econ_2[d_inc_econ_2$year %in% years_out_samp,])
out_samp_truth <- d_inc_econ_2$pv2p[d_inc_econ_2$year %in% years_out_samp]
mean(out_samp_pred - out_samp_truth)
```

```{r, echo=FALSE, warning = FALSE, message = FALSE}
out_samp_errors <- sapply(1:1000, function(i) {
  years_out_samp <- sample(d_inc_econ_2$year, 9) 
  mod <- lm(pv2p ~ GDP_growth_quarterly, 
            d_inc_econ_2[!(d_inc_econ_2$year %in% years_out_samp),])
  out_samp_pred <- predict(mod, d_inc_econ_2[d_inc_econ_2$year %in% years_out_samp,])
  out_samp_truth <- d_inc_econ_2$pv2p[d_inc_econ_2$year %in% years_out_samp]
  mean(out_samp_pred - out_samp_truth)
})

mean(abs(out_samp_errors))

hist(out_samp_errors,
     xlab = "",
     main = "Mean Out-of-Sample Residual\n(1000 Runs of Cross-Validation)")
```

```{r, echo=FALSE, warning = FALSE, message = FALSE}
GDP_new <- d_fred |> 
  filter(year == 2024 & quarter == 2) |> 
  select(GDP_growth_quarterly)
```

```{r, echo=FALSE, warning = FALSE, message = FALSE}
predict(reg_econ_2, GDP_new)
```

```{r, echo=FALSE, warning = FALSE, message = FALSE}
predict(reg_econ_2, GDP_new, interval = "prediction")
```


Sources:
1: https://www.pewresearch.org/politics/2024/09/09/issues-and-the-2024-election/
